---
title: "BarSeq Analysis - CRISPR Knockout Screen"
author: Jeffrey Lewis & Carson Stacy
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

last updated: `r Sys.Date()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed("1492")
```

# Description

This exercise introduces analysis for experiments that use a genome-wide
screen. This example is based on a CRISPR-based knockout screen in
yeast, but similar principles apply to other types of screens (e.g.,
RNAi, overexpression, Tn-Seq).

# Learning Outcomes

At the end of this exercise, you should be able to:

-   Describe the data generated when using genome-wide screen.
-   Describe the analysis steps of typical analysis workflows for
    genome-wide screens.
-   Utilize the edgeR package to identify genes with
-   Interpret biological significance of the results from a genome-wide
    screen.

# Getting Things Setup

As usual, make sure we have the right packages for this exercise

```{r ready-packages}
if (!require("pak")) install.packages("pak"); library(pak)

# let's load all of the files we were using and want to have again today
# pak(c("statmod", # required dependency, need to load manually on some macOS versions.
#     "purrr", # for working with lists (beautify column names)
#     "webshot2", # allow for pdf of output table.
#     "reactable") # for pretty tables.
# )

# We also need these Bioconductor packages today.
pak(c("edgeR", "AnnotationDbi", "org.Sc.sgd.db"))
```

```{r}
# Load the packages we need
library(edgeR)
library(AnnotationDbi)
library(org.Sc.sgd.db)
library(tidyverse)
# library(purrr)
# library(webshot2)
# library(reactable)
library(GenomicAlignments)
library(Rsubread)
library(clusterProfiler)
```

# Load the Data

The data we will be using is a CRISPR knockout screen in yeast. This
data is a subset of an upcoming publication from our lab. For analysis,
we need a count matrix of the number of reads that map to each guide
(and thus gene) in the yeast (pan)genome.

What do we need to analyze CRISPR screen data? - Sequencing (FASTQ or
BAM) files - Guide sequences (FASTA) files - Often need to convert a
text file to fasta format.

## Raw Sequencing Results

Any type of genome-wide screen will generate a large amount of raw
sequencing data in the form FASTQ files. Similar to RNAseq, these FASTQ
files need to be processed to generate a count matrix, which contains
the number of reads that map to each guide or gene in the screen.

We will work with **mapped BAM** files in this exercise for simplicity.

```{r}
# Download CRISPR guide sequences library
url_guides <- "https://github.com/clstacy/GenomicAnalysis/raw/main/Reference/CRISPR_gRNA_seqs.csv"

# Create directories for the reference and index files
if (!dir.exists("~/Desktop/Genomic_Data_Analysis/CRISPR_screen")) {
  dir.create("~/Desktop/Genomic_Data_Analysis/CRISPR_screen/reference",
             recursive = TRUE)
  dir.create("~/Desktop/Genomic_Data_Analysis/CRISPR_screen/index",
             recursive = TRUE)
}

# Download the guide sequences
download.file(url_guides, file.path("~/Desktop/Genomic_Data_Analysis/CRISPR_screen/reference", basename(url_guides)))

# load and create an index of the guide sequences for mapping
CRISPR_guides <- read_csv("~/Desktop/Genomic_Data_Analysis/CRISPR_screen/reference/CRISPR_gRNA_seqs.csv", col_names = FALSE)
colnames(CRISPR_guides) <- c("gRNA", "seq", "gene")

CRISPR_guides[101:102, ]
```

## Prepare Reference Guide Library for Indexing and Mapping

```{r}
# Convert the guide sequences to a DNAStringSet object
sgRNAs <- DNAStringSet(CRISPR_guides$seq)
names(sgRNAs) <- CRISPR_guides$gRNA
sgRNAs

# write the DNAStringSet object to a FASTA file
writeXStringSet(sgRNAs, "~/Desktop/Genomic_Data_Analysis/CRISPR_screen/index/sgRNAs.fa", 
                format = "fasta")
```

## Building the Index

```{r}
buildindex(basename = "~/Desktop/Genomic_Data_Analysis/CRISPR_screen/index",
           reference = "~/Desktop/Genomic_Data_Analysis/CRISPR_screen/index/sgRNAs.fa",
           indexSplit = FALSE)
```

## Download BAM files

For the sake of time, we start with BAM files that have already been
mapped to our **sgRNAs** reference.

**Download the BAM files from the course folder
`Data Files/CRISPR_screen/BAM/` (link:**
[OneDrive](https://uark-my.sharepoint.com/:f:/r/personal/clstacy_uark_edu/Documents/Genomic%20Analysis%20(BIOL5800V-CEMB5900V)/Data%20Files/CRISPR_screen/BAM?csf=1&web=1&e=cYwb0L))
**and move them to the folder shown below:**

```{r}
if (!dir.exists("~/Desktop/Genomic_Data_Analysis/CRISPR_screen/bam")) {
    dir.create("~/Desktop/Genomic_Data_Analysis/CRISPR_screen/bam")
}
```

## Quantify Guide Counts with Aligned Reads

```{r}
bam_files <- dir(
  path = "~/Desktop/Genomic_Data_Analysis/CRISPR_screen/bam",
  pattern = "*.bam$",
  full.names = TRUE
)

counts <- list()

# Read each BAM file and tally up guides.
  for (i in seq_along(bam_files)) {
    temp <- readGAlignments(bam_files[i])
    counts[[i]] <- data.frame(table(seqnames(temp)), row.names = "Var1")
  }

  CRISPR_screen_counts <- do.call(cbind, counts)
  colnames(CRISPR_screen_counts) <- c(
    "Lab_NaCl_Rep1", "Lab_NaCl_Rep2", "Lab_NaCl_Rep3",
    "Lab_NoStress_Rep1", "Lab_NoStress_Rep2", "Lab_NoStress_Rep3",
    "Oak_NaCl_Rep1", "Oak_NaCl_Rep2", "Oak_NaCl_Rep3",
    "Oak_NoStress_Rep1", "Oak_NoStress_Rep2", "Oak_NoStress_Rep3"
  )
```

# Prepare count data

```{r}
# If needed, load the count data from a saved CSV instead:
# CRISPR_screen_counts <- readr::read_csv("https://github.com/clstacy/GenomicAnalysis/raw/main/Data/CRISPR_screen/Counts/CRISPR_screen_counts.csv") |> tibble::column_to_rownames("...1")

design_matrix <- data.frame(
  sample = colnames(CRISPR_screen_counts),
  condition = c(
    rep(c("Lab_NaCl", "Lab_NoStress"), each = 3),
    rep(c("Oak_NaCl", "Oak_NoStress"), each = 3)
  ),
  replicate = rep(1:3, 4)
)

# Convert the rownames to an explicit column for clarity
CRISPR_screen_counts_df <- CRISPR_screen_counts %>%
  tibble::rownames_to_column("sgRNA") %>%
  mutate(
    gene = gsub("-[0-9]+$", "", sgRNA),
    .after = sgRNA
  ) %>%
  mutate(
    gene = stringr::str_remove(gene, "core-")
  )

# Create a DGEList from edgeR
y <- DGEList(
  counts = as.matrix(CRISPR_screen_counts),
  samples = design_matrix,
  genes = CRISPR_screen_counts_df$gene,
  group = design_matrix$condition
)
y
```

# Exploratory Data Analysis

Exploratory data analysis (EDA) helps us spot potential issues.

## Check distribution of Guide Data

An example approach is to look at the total number of nonzero guides per
sample.

```{r}
# Let's quickly check how many guides are detected in each sample:
non_zero_guides <- colSums(y$counts > 0)

# Create a small data frame for plotting
df_nonzero <- tibble(
  Sample = names(non_zero_guides),
  Guides = as.integer(non_zero_guides)
)

# Plot with ggplot2
ggplot(df_nonzero, aes(x = Sample, y = Guides)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Number of Detected Guides per Sample",
    x = "Sample",
    y = "Number of Guides"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

## Check uniformity across guides

The **Gini index** is a measure of inequality in a distribution, first
developed in economics, it has proven very useful for detecting
uniformity of guides in a screen.

A Gini index of 0 indicates perfect equality, while a Gini index of 1
indicates perfect inequality.

In the context of guide counts, a low Gini index indicates that the
guides are evenly distributed across samples, while a high Gini index
indicates that some guides are much more abundant than others.

```{r}
gini(y$counts)

boxplot(gini(y$counts) ~ rownames(y$samples), ylab = "Gini Indices", xlab = "",
    cex.axis = 0.8)


guide_gini <- apply(y$counts, 2, gini)

gini_df <- tibble(
  Sample = colnames(y$counts),
  Gini = guide_gini
)


ggplot(gini_df, aes(x = Sample, y = Gini)) +
  geom_col(fill = "pink") +
  labs(
    title = "Gini Index of Guide Distributions",
    x = "Sample",
    y = "Gini Index"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Data Preprocessing

Guides lacking an appreciable numbber of reads in any sample should
normally be excluded from downstream analyses. There are multiple
rationales for doing so. - Biologically, guides that are not observed at
a biologically relevant level in any condition are deemed uninteresting
and are consequently disregarded. - From a statistical standpoint, the
elimination of guides with low counts enhances the reliability of
estimating the mean-variance relationship in the data and reduces the
number of statistical tests required for subsequent analyses focused on
differential abundance.

## Filtering

For filtering, guides with low expression were filtered out, retaining
as many guides as possible with meaningful counts. This was achieved
using the filterByExpr() function in the edgeR package, a tool commonly
applied in RNA-seq experiments for automated gene filtering. Initially,
the count matrix contained 29089 guides. After applying `filterByExpr()`,
6914 guides (24%%) were removed due to their low presence, leaving
22,175 guides for further analysis.

```{r}
# Filter out low abundance guides
keep.exprs <- filterByExpr(y, group = y$samples$group)
keep.exprs[1:10]  # Just a quick peek

y.filtered <- y[keep.exprs, , keep.lib.sizes = FALSE]

table(keep.exprs)  # How many guides remain?

cat("Number of guides after filtering:\n", nrow(y.filtered$counts))
```

```{r}
# How many unique genes?
unique_genes <- unique(y.filtered$genes)[[1]]
length(unique_genes)
```

It's worth noting that `filterByExpr()` often retains only one guide per
gene, which is sub optimal for CRISPR screens, where the collective
behavior of multiple guides targeting the same gene is essential for
validation. This screen has high abundance of most guide, which is often
not the case.

```{r}
# Quick look at how many sgRNAs remain per gene:
nsgrna <- table(table(y.filtered$genes))

# Convert it to a data frame for plotting
dist_df <- tibble(
  sgRNA_count = as.factor(names(nsgrna)),
  freq = as.numeric(nsgrna)
)

ggplot(dist_df, aes(x = sgRNA_count, y = freq)) +
  geom_col(fill = "#FF9999") +
  labs(
    title = "Distribution of # of sgRNAs per gene (After Filtering)",
    x = "# of sgRNAs",
    y = "Count of Genes"
  ) +
  theme_classic() 
```

```{r}
# We can check the distribution of counts before and after filtering:
cols <- c("red", "orange", "blue", "skyblue")

nsamples <- ncol(y)
col.density <- RColorBrewer::brewer.pal(nsamples, "Paired")

# Raw
cpm_raw <- cpm(y, log = TRUE)
plot(density(cpm_raw[, 1], bw = 0.4), col = cols[factor(y$samples$group)],
     lwd = 2,
     ylim = c(0, 1), las = 2, main = "A. Raw Data", xlab = "logCPM")
for (i in 2:nsamples) {
  den <- density(cpm_raw[, i], bw = 0.4)
  lines(den$x, den$y, col = cols[factor(y$samples$group)][i],
        lwd = 2)
}
legend("topright", colnames(y$counts), text.col = cols[factor(y$samples$group)],
       bty = "n", cex = 0.6)

# Filtered
nsamples_filt <- ncol(y.filtered)
cpm_filt <- cpm(y.filtered, log = TRUE)
plot(density(cpm_filt[, 1], bw = 0.4), col = cols[factor(y$samples$group)][1], lwd = 2,
     ylim = c(0, 1), las = 2, main = "B. Filtered Data", xlab = "logCPM")
for (i in 2:nsamples_filt) {
  den <- density(cpm_filt[, i], bw = 0.4)
  lines(den$x, den$y, col = cols[factor(y$samples$group)][i], lwd = 2)
}
legend("topright", colnames(y.filtered$counts), text.col = cols[factor(y$samples$group)], bty = "n", cex = 0.6)
```

```{r}
# TMM Normalize
yy <- y.filtered[, order(y.filtered$samples$group, colSums(y.filtered$counts))]

yy$samples$norm.factors <- normLibSizes(yy$counts, method = "TMM")

yy$samples
```

```{r}
lcpm <- cpm(yy, log = TRUE)
cols <- c("red", "orange", "blue", "lightblue")
plotMDS(lcpm,
  col = cols[factor(yy$samples$group)],
  pch = 16,
  cex = 1,
  main = "MDS Plot of TMM-Normalized Data"
)
legend("bottomleft",
  legend = levels(factor(yy$samples$group)),
  col = cols, pch = 16, cex = 0.8
)
```

Check the normalized data using MD plots

```{r}
par(mfrow = c(1,3))
for (i in 1:ncol(yy$counts)) {
  plotMD(yy, column = i)
  abline(h = 0, col = "red", lty = 2, lwd = 2)
}
```

Ideally, we expect the majority of the counts to be
concentrated near the zero log-fold change line. However, this is rarely
true for CRISPR experiments due to the sparse nature of CRISPR screen
counts. The diagonal lines in the lower-left portion of the plot
correspond to guides with counts of very low numbers. And it is typical
for treated samples to observe a positive skew, with a number of very
highly upregulated guides in the upper right of the plots.

# Differential Abundance Analysis (Guides)

## Create Design Matrix

```{r}
yy$samples$group <- factor(yy$samples$group, levels = c("Lab_NoStress", "Lab_NaCl", "Oak_NoStress", "Oak_NaCl"))
design <- model.matrix(~0 + group, data = yy$samples)
design
```

## Estimate Dispersion

```{r}
yy.norm <- estimateDisp(yy, design)
yy.norm$common.dispersion

# Plot the dispersion estimates
plotBCV(yy.norm)

# Show biological coefficient of variation
bcv <- sqrt(yy.norm$common.dispersion)
bcv
```

A BCV below 0.3 is quite typical for well-controlled samples of an
isogenic background. The trended dispersion shows that when the
abundance gets higher, the variation decreases. At low logCPM, the
dispersion is very large.

## Fit the model

```{r}
fit <- glmQLFit(yy.norm, design, robust = TRUE)
plotQLDisp(fit)
```

## Create contrast matrix

```{r}
my.contrasts <- makeContrasts(
  Lab_SaltvsNone = (groupLab_NaCl - groupLab_NoStress),
  Oak_SaltvsNone = (groupOak_NaCl - groupOak_NoStress),
  LabvsOak_SaltvsNone= ( (groupOak_NaCl - groupLab_NaCl) - 
                           (groupOak_NoStress - groupLab_NoStress) ),
  levels = design
)

my.contrasts
```

## Helper Functions

These are used below for analyzing and plotting each contrast

```{r}
# 1) Convert topTags output to a consistent format.
format_topTags <- function(qlf_result, comparison_name) {
  df <- topTags(qlf_result, n = Inf, sort.by = "PValue")$table %>%
    mutate(comparison = comparison_name) %>%
    rename( log2FoldChange = logFC, pvalue  = PValue,
      gene_symbol = genes, adjusted_pvalue= FDR
    ) %>%
    drop_na()
  return(df)
}

# 2) Create a generic volcano plot.
volcano_plot <- function(df, title) {
  # Label significance
  df <- df %>% mutate(diffabundant = case_when(
    adjusted_pvalue > 0.05  ~ "Not Significant",
    (log2FoldChange >= 1) & adjusted_pvalue <= 0.05 ~ "Up",
    (log2FoldChange <= -1) & adjusted_pvalue <= 0.05 ~ "Down",
    TRUE ~ "Pass p-value cutoff"))

  # Add gene names for top hits
  # We pick top 20 up and top 20 down by absolute fold change.
  top_genes <- df %>%
    dplyr::filter(diffabundant %in% c("Up", "Down")) %>%
    group_by(diffabundant) %>%
    slice_max(order_by = abs(log2FoldChange), n = 20)

  df$delabel <- ifelse(df$gene_symbol %in% top_genes$gene_symbol, df$GENENAME, NA)

  ggplot(data = df,aes(x = log2FoldChange, y = -log10(adjusted_pvalue),
      col = diffabundant, label = delabel )) +
    geom_point() +
    theme_minimal() +
    ggrepel::geom_text_repel(max.overlaps = Inf, size = 2) +
    scale_color_manual(values = c( "Down" = "blue", "Up" = "red",
      "Pass p-value cutoff" = "gray", "Not Significant" = "black")) +
    geom_vline(xintercept = c(-1, 1), color = "darkred") +
    geom_hline(yintercept = -log10(0.05), color = "darkred") +
    labs(title = title, color = "Abundance Change",
      x = "log2 Fold Change", y = "-log10(Adjusted p-value)" )
}

# 3) A convenience function to run the test, reformat, annotate, and plot.
run_contrast_and_plot <- function(fit, contrast_matrix, contrast_name, plot_title) {
  # Run QLF-test
  qlf <- glmQLFTest(fit, contrast = contrast_matrix)
  # Format topTags results
  df <- format_topTags(qlf, contrast_name)
  # Annotate with gene info from org.Sc.sgd.db
  df <- df %>%
    left_join(
      AnnotationDbi::select(
        org.Sc.sgd.db,
        keys = df$gene_symbol,
        columns = c("GENENAME", "DESCRIPTION")
      ), by = c("gene_symbol" = "ORF") ) %>%
    distinct()
  
  # Volcano plot
  vplot <- volcano_plot(df, plot_title)
  return(list(results = df, plot = vplot))
}
```

# Test for differential Guide Abundance

## 1.) Lab strain NaCl vs No Stress

Let's look at the Lab strain salt vs control, to see which guides show
differential abundance.

```{r, warning=FALSE, message=F}
lab_salt <- run_contrast_and_plot(
  fit,
  my.contrasts[, "Lab_SaltvsNone"],
  "Lab_SaltvsNone",
  "Lab Strain: Salt vs No Stress"
)

lab_salt$plot  # display volcano plot
head(lab_salt$results)
```

## 2.) Oak strain NaCl vs No Stress

Same analysis, but for the Oak isolate of yeast
```{r, warning=FALSE, message=F}
oak_salt <- run_contrast_and_plot(
  fit,
  my.contrasts[, "Oak_SaltvsNone"],
  "Oak_SaltvsNone",
  "Oak Strain: Salt vs No Stress"
)

oak_salt$plot
head(oak_salt$results)
```

## 3.) Lab vs Oak: Difference in Salt Effect Between Strains

We can also look for guide differences for salt stress when we compare
difference in strain responses to stress.

```{r, warning=FALSE, message=F}
oak_vs_lab <- run_contrast_and_plot(
  fit,
  my.contrasts[, "LabvsOak_SaltvsNone"],
  "LabvsOak_SaltvsNone",
  "Lab vs Oak Strain: Salt vs No Stress"
)

oak_vs_lab$plot
head(oak_vs_lab$results)
```

## Save Outputs to Text Files
```{r}
if (!dir.exists("~/Desktop/Genomic_Data_Analysis/CRISPR_screen/results")) {
  dir.create("~/Desktop/Genomic_Data_Analysis/CRISPR_screen/results",
             recursive = FALSE)
}

write.csv(lab_salt$results, 
          file = "~/Desktop/Genomic_Data_Analysis/CRISPR_screen/results/Lab_salt_response.csv",     
          row.names = TRUE)

write.csv(oak_salt$results, 
          file = "~/Desktop/Genomic_Data_Analysis/CRISPR_screen/results/Oak_salt_response.csv",     
          row.names = TRUE)

write.csv(oak_vs_lab$results, 
          file = "~/Desktop/Genomic_Data_Analysis/CRISPR_screen/results/Lab_vs_Oak_salt_response.csv", 
          row.names = TRUE)
```

# Differential Abundance Analysis (Gene-Level)

The above analysis is very similar to what we have done before with
edgeR, but the differential abudance was measured **at the guide
level**, rather than the gene level, which is often the level of
interest.

The edgeR package includes a function called `fry()` that can be used to
perform differential abundance analysis at the **gene level**. This
function takes the guide-level results and aggregates them to the gene
level, allowing us to identify genes that are differentially abundant
across samples. Note it only tests for significance of direction, and no
estimation of fold change is performed.

```{r}
# Create a list of genes with sgRNA indices for the 'fry' function
genesymbols <- yy.norm$genes[, 1]

genesymbollist <- list()
unq <- unique(genesymbols)
unq <- unq[!is.na(unq)]

for (i in unq) {
  sel <- (genesymbols == i & !is.na(genesymbols))
  # example threshold: at least 3 guides per gene to test
  if (sum(sel) >= 3) {
    genesymbollist[[i]] <- which(sel)
  }
}

# fry results for Lab strain
fryLab_SaltvsNone <- fry(
  y     = yy.norm,
  index = genesymbollist,
  design = design,
  contrast = my.contrasts[, "Lab_SaltvsNone"]
) |>
  rownames_to_column("gene_symbol") %>%
  left_join(
      AnnotationDbi::select(
        org.Sc.sgd.db,
        keys = .$gene_symbol,
        columns = c("GENENAME", "DESCRIPTION")
      ),
      by = c("gene_symbol" = "ORF")
    ) %>%
    distinct()

View(fryLab_SaltvsNone)
```

```{r}
# fry results for Oak strain
fryOak_SaltvsNone <- fry(
  y     = yy.norm,
  index = genesymbollist,
  design = design,
  contrast = my.contrasts[, "Oak_SaltvsNone"]
) |>
  rownames_to_column("gene_symbol") %>%
  left_join(
      AnnotationDbi::select(
        org.Sc.sgd.db,
        keys = .$gene_symbol,
        columns = c("GENENAME", "DESCRIPTION")
      ),
      by = c("gene_symbol" = "ORF")
    ) %>%
    distinct()
```

```{r}
# fry results for Lab-vs-Oak contrast
fryLabvsOak_SaltvsNone <- fry(
  y     = yy.norm,
  index = genesymbollist,
  design = design,
  contrast = my.contrasts[, "LabvsOak_SaltvsNone"]
) |>
  rownames_to_column("gene_symbol") %>%
  left_join(
      AnnotationDbi::select(
        org.Sc.sgd.db,
        keys = .$gene_symbol,
        columns = c("GENENAME", "DESCRIPTION")
      ),
      by = c("gene_symbol" = "ORF")
    ) %>%
    distinct()

head(fryLabvsOak_SaltvsNone)
```

# Functional Enrichment Analysis

We can perform functional enrichment analysis on the differentially
abundant genes using the `clusterProfiler` package, as we have done
before.

```{r}
# Create a list of DE genes for each contrast
DE_genes_list <- list(
  LabSalt = lab_salt$results %>%
    dplyr::filter(adjusted_pvalue < 0.05) %>%
    pull(gene_symbol),
  OakSalt = oak_salt$results %>%
    dplyr::filter(adjusted_pvalue < 0.05) %>%
    pull(gene_symbol),
  LabvsOak = oak_vs_lab$results %>%
    dplyr::filter(adjusted_pvalue < 0.05) %>%
    pull(gene_symbol)
)

# Use compareCluster to run GO enrichment on each contrast.
GO_CRISPR_results <- clusterProfiler::compareCluster(
  geneCluster = DE_genes_list,
  fun = "enrichGO",
  keyType = "ORF",
  OrgDb = org.Sc.sgd.db,
  ont   = "BP"
)

GO_CRISPR_results |>
  as_tibble() |>
  View()
```

## Plot GO Enrichment Results

```{r}
dotplot(GO_CRISPR_results)
```

```{r, fig.height=8, fig.width=8}
GO_CRISPR_results %>%
  mutate(richFactor = Count / as.numeric(sub("/\\d+", "", BgRatio))) %>%
  ggplot(.,
         showCategory = 6,
         aes(richFactor, fct_reorder(Description, richFactor))) +
    geom_segment(aes(xend = 0, yend = Description)) +
    geom_point(aes(color = p.adjust, size = Count)) +
    scale_color_gradientn(
      colours = c("#f7ca64", "#46bac2", "#7e62a3"),
      trans = "log10",
      guide = guide_colorbar(reverse = TRUE, order = 1)
    ) +
    scale_size_continuous(range = c(2, 10)) +
    xlab("Rich Factor") +
    ylab(NULL) +
    ggtitle("Biological Processes") +
    facet_wrap(~Cluster, ncol=2) +
    theme_bw()
```

```{r}
ggVennDiagram::ggVennDiagram(DE_genes_list
              ) +
  scale_x_continuous(expand = expansion(mult = .2)) +
  scale_fill_distiller(palette = "RdBu"
  )
```

## GO Enrichment of Missing Guides

Which biological processes are enriched for guides that were excluded from the analysis?
```{r}
# GO enrichment of missing guides
y.missing <- y[!keep.exprs, , keep.lib.sizes = FALSE]

GO_missing_genes <- y.missing$genes %>%
  unique() %>%
  pull(genes) %>%
  clusterProfiler::enrichGO(
    OrgDb = org.Sc.sgd.db,
    keyType = "ORF",
    ont = "BP",
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05
  ) %>%
  mutate(richFactor = Count / as.numeric(sub("/\\d+", "", BgRatio)))

GO_missing_genes %>%
  ggplot(.,
         showCategory = 10,
         aes(richFactor, fct_reorder(Description, richFactor))) +
    geom_segment(aes(xend = 0, yend = Description)) +
    geom_point(aes(color = p.adjust, size = Count)) +
    scale_color_gradientn(
      colours = c("#f7ca64", "#46bac2", "#7e62a3"),
      trans = "log10",
      guide = guide_colorbar(reverse = TRUE, order = 1)
    ) +
    scale_size_continuous(range = c(2, 10)) +
    xlab("Rich Factor") +
    ylab(NULL) +
    ggtitle("Biological Processes") +
    theme_bw()
```

System information for reproducibility:

```{r}
pander::pander(sessionInfo())
```
