---
title: "Read Mapping"
author: "Carson Stacy & Jeffrey Lewis"
date: "Fall 2023"
output: html_document
---

last updated: `r Sys.Date()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed("1492")
```

As usual, make sure we have the right packages for this exercise

```{r ready-packages}
if (!require("pacman")) install.packages("pacman"); library(pacman)

# let's load all of the files we were using and want to have again today
p_load("tidyverse", "knitr", "readr",
       "pander", "BiocManager", 
       "dplyr", "stringr")

# We also need the Bioconductor packages "Rsubread" for today's activity.
p_load("Rsubread")

```

Previously, we filtered and trimmed our raw fastq files. They should be in the folder below, unless you chose a different place to store them.

```{r get-files}

dir_trimmed.fq_files <- "~/Desktop/Genomic_Data_Analysis/Data/Trimmed_rfastp"

trimmed_fastq_files <- list.files(path = dir_trimmed.fq_files, 
                                  pattern = ".fastq.gz$", 
                                  full.names = TRUE)
trimmed_fastq_files
```

You should see the full paths to all 16 trimmed fastq files that we will be mapping to the reference genome today. If not, check your computer for those files and either change the file path or re-download from "03_Working_with_Sequences.Rmd"

# Alignment

Read sequences are stored in compressed (gzipped) FASTQ files. Before the differential expression analysis can proceed, these reads must be aligned to the yeast genome (or transcriptome), and then reads counts must be assigned for each transcript (gene). This can be achieved with functions in the Rsubread package. For today, we will just perform the mapping portion.

## Retrieve the genome

We will use a bash code chunk to download the latest *Saccharomyces* cerevisiae genome. The genome is in FASTA format, with the chromosome names as headers and then the entire DNA sequence of each chromosome following (so like a FASTQ file without the quality scores, and "only" 17 reads--16 autosomes and the mitochondrial genome). Genomes can be retrieved from a variety of sources, but if your genome is on the Ensembl database, that is often the most up-to-date. Genomes can also be found on NCBI or organism-specific databases.

```{bash fetch-genome}
# Define the destination file path
# You can change this file path to the path you want your data to go, or leave it.
REF_DIR="/Users/$USER/Desktop/Genomic_Data_Analysis/Reference"

# make that directory if it doesn't already exist
mkdir -p $REF_DIR

# Define the URL of reference genome
# (latest from ensembl)
url="ftp://ftp.ensembl.org/pub/release-110/fasta/saccharomyces_cerevisiae/dna/Saccharomyces_cerevisiae.R64-1-1.dna.toplevel.fa.gz"


# Check if the file already exists at the destination location
if [ ! -f "$REF_DIR/Saccharomyces_cerevisiae.R64-1-1.dna.toplevel.fa.gz" ]; then
    echo "Reference genome not found, downloading..."
    # If the file does not exist, download it using curl
    curl -o "$REF_DIR/Saccharomyces_cerevisiae.R64-1-1.dna.toplevel.fa.gz" "$url"
    echo "Downloading finished"
else
    echo "File already exists at $REF_DIR Skipping download."
fi
```

## Build the index

The first step in performing the alignment is to build an index (which makes searching for alignments much, much faster). To build an index you need the FASTA file (.fa) of the genome. For larger eukaryotic genomes, this step could take 30 - 60 minutes. Because yeast has a relatively small genome size (12 Mb), we are able to build the full index in class in a few minutes.

```{r build-index}
library(Rsubread)

# Set path of the reference fasta file
reference_genome = path.expand("~/Desktop/Genomic_Data_Analysis/Reference/Saccharomyces_cerevisiae.R64-1-1.dna.toplevel.fa.gz")

index_reference_genome = path.expand("~/Desktop/Genomic_Data_Analysis/Reference/index_rsubread_Saccharomyces_cerevisiae.R64-1-1")

# build the index
buildindex(basename=index_reference_genome, reference=reference_genome)
```

The next step after indexing is to actually align our reads to the genome. We can see the arguments available with the align function from the Rsubread package

```{r get-args}
args(align)
```

This process takes some time to finish, and is generally the rate-limiting step for any sequencing bioinformatic pipeline. To see what the arguments actually mean, type ?align into the Console.

```{r align-reads}
# run the alignment on all of the trimmed_fastq_files
align(index=index_reference_genome, 
      readfile1=trimmed_fastq_files,
      type = "rna",
      input_format = "gzFASTQ",
      output_format = "BAM",
      unique = TRUE,
      nBestLocations = 1,
      nthreads=4
      )
```

The output for each alignment is a binary alignment map (bam) file. There will be one bam file generated from each FASTQ file whose reads are aligned to the genome.

We can get a summary of the proportion of reads that mapped to the reference genome using the propmapped function. Depending on the organism and read type (single vs paired end), 70-90% of the reads should align to the genome.

```{r output-summary}
# create an object in R listing
bam_files <- list.files(path = dir_trimmed.fq_files, pattern = ".BAM$", full.names = TRUE)
bam_files

# find the proportion of reads that mapped for each sample
props <- propmapped(files=bam_files)

props %>% print()
```

# Pseudomapping with Salmon

Salmon is a widely used "pseudoaligner," which means that instead of looking for matches in the genome, the aligner basically determines which gene the read came from, but not exactly where in the gene in aligns (hence the pseudoaligner terminology). This saves a ton of time, and for RNA-seq, we just care which reads align to each gene for counting purposes, not the precise location where the read maps within the gene.

. It is not available to use in R, but we can use bash code chunks to run it in the same markdown document.

First, we need to create a new conda environment for salmon.

**Depending on your computer, we might need to run this code in terminal.**

```{bash create-condaEnvironment, engine.opts='-l'}
## Warning, if you did not complete Working_with_Sequences.Rmd activity, 
#    your conda might not be set up correctly for this code.

# create an environment for our pseudomapping with Salmon
# this code is "extra" because it only creates env if not already existing.
if conda info --envs | grep -q salmon; then echo "environment 'salmon' already exists"; else CONDA_SUBDIR=osx-64 conda create -y -n salmon -c conda-forge -c bioconda salmon=1.10.0; fi
# the channel priority order above is needed to get a recent version via conda.

# see available conda environments
conda env list

# activate our QC environment
conda activate salmon

# make sure desired packages are working
which salmon

# help page for using salmon
salmon -h

# it's always good coding practice to deactivate 
# a conda environment at the end of a chunk
conda deactivate
```

## Download transcriptome

To make an index for Salmon, we need transcript sequences in the FASTA format.

```{bash fetch-transcriptome}
# Define the destination file path
# Be sure to change this file path to the path you want your data to go
REF_DIR="/Users/$USER/Desktop/Genomic_Data_Analysis/Reference"

# make that directory if it doesn't already
mkdir -p $REF_DIR

# Define the URL of reference transcriptome
# (latest from ensembl)
url="ftp://ftp.ensembl.org/pub/release-110/fasta/saccharomyces_cerevisiae/cdna/Saccharomyces_cerevisiae.R64-1-1.cdna.all.fa.gz"


# Check if the file already exists at the destination location
if [ ! -f "$REF_DIR/Saccharomyces_cerevisiae.R64-1-1.cdna.all.fa.gz" ]; then
    echo "Reference transcriptome not found, downloading..."
    # If the file does not exist, download it using curl
    curl -o "$REF_DIR/Saccharomyces_cerevisiae.R64-1-1.cdna.all.fa.gz" "$url"
    echo "Downloading finished"
else
    echo "File already exists at $REF_DIR Skipping download."
fi
```

## Building the Salmon index

Salmon can index by using the command `salmon index`. A recent feature update to Salmon includes an option to map to "decoys," which are DNA sequences that should not be found in the FASTQ files (e.g., non-coding DNA sequences should not be found in an RNA-seq library). Using decoys allows Salmon to identify reads that (pseudo)align better to the genome instead of the transcriptome, and thus are likely to have mapping errors. We will use the entire genome as the decoy for our index, because the *S. cerevesiae* genome is small. You can read more at: <https://salmon.readthedocs.io/en/latest/salmon.html#preparing-transcriptome-indices-mapping-based-mode>.

```{bash index-salmon, engine.opts='-l', message=FALSE}
# We need to set a variable for where the transcriptome file is
REF_DIR="/Users/$USER/Desktop/Genomic_Data_Analysis/Reference"
TRANSCRIPTOME="/Users/$USER/Desktop/Genomic_Data_Analysis/Reference/Saccharomyces_cerevisiae.R64-1-1.cdna.all.fa.gz"
GENOME="/Users/$USER/Desktop/Genomic_Data_Analysis/Reference/Saccharomyces_cerevisiae.R64-1-1.dna.toplevel.fa.gz"

# Activate conda salmon environment
conda activate salmon

# Run a script that generates a decoy.txt file from the genome we downloaded
grep "^>" <(gunzip -c $GENOME) | cut -d " " -f 1 > $REF_DIR/decoys.txt
sed -i.bak -e 's/>//g' $REF_DIR/decoys.txt

# Combine the transcriptome and genome into a single file for indexing
cat $TRANSCRIPTOME $GENOME > $REF_DIR/gentrome.fasta.gz


# We will use the yeast, but it needs to be indexed by salmon
salmon index -t $REF_DIR/gentrome.fasta.gz -d $REF_DIR/decoys.txt -p 4 -i $REF_DIR/index_salmon_Saccharomyces_cerevisiae.R64-1-1

conda deactivate
```

Notice that we combined the fasta file of the transcriptome with the fasta file of the entire genome (in that order) into the gentrome.fasta.gz file which was then indexed.

Salmon is a pseudomapper, so it doesn't create sam/bam files and is instead able to count directly from the FASTQ files. We will do the pseudoaligning and counting all in one step in the next activity.

## Questions

With Rsubread:

Question 1: Try aligning the FASTQ files allowing multi-mapping reads (set unique = FALSE) and allowing for up to 6 "best" locations to be reported (nBestLocations = 6). Specify the output file names (bam_files_multi) by substituting ".fastq.gz" with ".multi.bam" so we don't overwrite our unique alignment bam files.

```{r multimap-align, eval=FALSE}
# Define the pattern and replacement
pattern <- "\\.fastq\\.gz$"
replacement <- "subread.multi.bam"

# Create the new file names
bam_files_multi <- gsub(pattern, replacement, trimmed_fastq_files)

# update this code to run with Rsubread multimapping, as described above.
align(index=index_reference_genome, 
      readfile1=trimmed_fastq_files,
      output_file = ___________,
      type = "rna",
      input_format = "gzFASTQ",
      output_format = "BAM",
      unique = ____,
      nBestLocations = ____,
      nthreads=6
      )
```

Question 2: Look at the proportion of reads mapped and see if we get any more reads mapping by specifying a less stringent criteria.

With Salmon:

Question 3: What are the pros and cons of using Salmon vs Rsubread for mapping reads?

Be sure to knit this file into a pdf or html file once you're finished.

System information for reproducibility:

```{r}
pander::pander(sessionInfo())
```
